---
title: "GroupM Analysis Notebook MARS"
author: "GroupM"
date: "2025-02-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This R notebook details the implementation of MARS on the 3 different datasets, using the earth package to set up the models.

## Data Import and Cleaning

```{r}
# Install packages if not already installed
if (!requireNamespace("Metrics", quietly = TRUE)) install.packages("Metrics", quiet = TRUE)
if (!requireNamespace("earth", quietly = TRUE)) install.packages("earth", quiet = TRUE)
if (!requireNamespace("MASS", quietly = TRUE)) install.packages("MASS", quiet = TRUE)

# Load packages with suppressed warnings
suppressWarnings({
  suppressPackageStartupMessages({
    library(MASS)    # Load the MASS package for robust linear models
    library(earth)   # Load earth package for MARS
    library(Metrics) # Load Metrics package for MAE, RMSE, MAPE
  })
})
```

```{r}
# Load the CleanDataset.csv file
clean_dataset <- read.csv('data/Clean_Dataset.csv')
clean_dataset <- clean_dataset[, !(names(clean_dataset) %in% c("X", "flight"))]

allclasses_dataset <- clean_dataset
economy_dataset <- clean_dataset[clean_dataset$class == "Economy", ]
business_dataset <- clean_dataset[clean_dataset$class == "Business", ]
```

```{r}
# Scale the values of the columns duration and days_left
allclasses_dataset$duration <- scale(allclasses_dataset$duration)
allclasses_dataset$days_left <- scale(allclasses_dataset$days_left)
economy_dataset$duration <- scale(economy_dataset$duration)
economy_dataset$days_left <- scale(economy_dataset$days_left)
business_dataset$duration <- scale(business_dataset$duration)
business_dataset$days_left <- scale(business_dataset$days_left)
```

```{r}
# Use log transformation for price
allclasses_dataset$price <- log(allclasses_dataset$price)
economy_dataset$price <- log(economy_dataset$price)
business_dataset$price <- log(business_dataset$price)
```

```{r}
# Remove the class column for the separate models
economy_dataset$class <- NULL
business_dataset$class <- NULL
```

```{r}
# Show the dimensions and the first rows of the datasets
cat("\nDimensions of the full dataset (all classes):\n")
dim(allclasses_dataset)
cat("\nFirst rows of the full dataset (all classes):\n")
head(allclasses_dataset)
cat("\nDimensions of the economy class dataset:\n")
dim(economy_dataset)
cat("\nDimensions of the business class dataset:\n")
dim(business_dataset)
cat("\nFirst rows of the economy class dataset:\n")
head(economy_dataset)
cat("\nFirst rows of the business class dataset:\n")
head(business_dataset)
```

```{r}
set.seed(123)  # For reproducibility
```

```{r}
# Create train and test sets for allclasses dataset
train_indices_allclasses <- sample(1:nrow(allclasses_dataset),
                                   size = 0.8 * nrow(allclasses_dataset))
train_allclasses <- allclasses_dataset[train_indices_allclasses, ]
test_allclasses <- allclasses_dataset[-train_indices_allclasses, ]
```

```{r}
# Create train and test sets for economy dataset
train_indices_economy <- sample(1:nrow(economy_dataset),
                                size = 0.8 * nrow(economy_dataset))
train_economy <- economy_dataset[train_indices_economy, ]
test_economy <- economy_dataset[-train_indices_economy, ]
```

```{r}
# Create train and test sets for business dataset
train_indices_business <- sample(1:nrow(business_dataset),
                                 size = 0.8 * nrow(business_dataset))
train_business <- business_dataset[train_indices_business, ]
test_business <- business_dataset[-train_indices_business, ]
```

```{r}
# Show the dimensions and the first rows of the train and test sets
cat("Dimensions of training set for all classes (80% of data):\n")
dim(train_allclasses)
```
```{r}
cat("\nDimensions of test set for all classes (20% of data):\n") 
dim(test_allclasses)
```
```{r}
cat("\nDimensions of training set for economy class (80% of data):\n")
dim(train_economy)
```
```{r}
cat("\nDimensions of test set for economy class (20% of data):\n")
dim(test_economy)
```
```{r}
cat("\nDimensions of training set for business class (80% of data):\n")
dim(train_business)
```
```{r}
cat("\nDimensions of test set for business class (20% of data):\n")
dim(test_business)
```
## MARS - Complete model

```{r}
# Construct the formula for the model with grade 2 numerical variables
formula <- as.formula(paste("price ~", 
                            paste(c(names(train_allclasses)[-which(names(train_allclasses) == "price")], 
    "I(duration^2)", "I(days_left^2)"), collapse = " + ")))
cat("Constructed formula for the model:\n")
print(formula)
```
Train the model using the internal algorithm to automatically choose terms and predictors:

```{r}
# Create MARS model for allclasses
mars_model_allclasses <- earth(formula,
                             data = train_allclasses,
                             degree = 4,  # Fixed degree
                            )
# Print model results
summary(mars_model_allclasses)

par(mfrow = c(2, 2))  # Set up the plotting area
plot(mars_model_allclasses)
```
We will use the evimp function from the 'earth' package to obtain the variable importance from the MARS model.

```{r, fig.width=10, fig.height=6}
evimp <- evimp(mars_model_allclasses, trim=TRUE, sqrt.=TRUE)
plot(evimp)
```
The plot illustrates the importance of the variables in the MARS model, highlighting that the most important predictor is the class, followed by the two numerical variables of degree 1.

```{r}
# Make predictions first
test_allclasses$duration_squared <- test_allclasses$duration^2
test_allclasses$days_left_squared <- test_allclasses$days_left^2
predictions_allclasses <- predict(mars_model_allclasses, newdata = test_allclasses)

# Calculate the residuals
residuals_mars_allclasses <- test_allclasses$price - predictions_allclasses

# Calculate the log-likelihood (fixed array conformability issue)
log_likelihood <- -0.5 * length(residuals_mars_allclasses) * (log(2 * pi) +
                  log(var(residuals_mars_allclasses))) - 
                  sum(residuals_mars_allclasses^2) / (2 * var(residuals_mars_allclasses))

# Calculate AIC and BIC
n <- length(residuals_mars_allclasses)  # Number of observations
k <- length(mars_model_allclasses$coefficients)  # Number of parameters

aic_value <- -2 * log_likelihood + 2 * k
bic_value <- -2 * log_likelihood + log(n) * k

# Print the results
cat("allclasses Model Performance:\n")
cat("AIC:", aic_value, "\n")
cat("BIC:", bic_value, "\n")

# Performance metrics for allclasses model using test set
# R^2 for log-transformed data
r2_allclasses_log <- 1 - sum((residuals_mars_allclasses)^2) / 
  sum((test_allclasses$price - mean(test_allclasses$price))^2)

# R^2 for original scale data
r2_allclasses_orig <- 1 - sum((exp(test_allclasses$price) - exp(predictions_allclasses))^2) / 
                      sum((exp(test_allclasses$price) - mean(exp(test_allclasses$price)))^2)

mae_allclasses <- mae(exp(test_allclasses$price), exp(predictions_allclasses))  
rmse_allclasses <- rmse(exp(test_allclasses$price), exp(predictions_allclasses))  
mape_allclasses <- mape(exp(test_allclasses$price), exp(predictions_allclasses)) 

cat("R^2 (log scale):", r2_allclasses_log, "\n")
cat("R^2 (original scale):", r2_allclasses_orig, "\n")
cat("MAE:", mae_allclasses, "\n")
cat("RMSE:", rmse_allclasses, "\n")
cat("MAPE:", mape_allclasses * 100, "%\n")
```
```{r}
# Create a histogram of the residuals
hist(residuals_mars_allclasses, breaks = 30, main = "Histogram of Residuals",
     xlab = "Residuals", ylab = "Frequency", col = "lightblue", border = "black", freq = FALSE)

# Add a density curve
lines(density(residuals_mars_allclasses), col = "red", lwd = 2)
```

## MARS - Economy model

```{r}
# Construct the formula for the model, excluding the 'class' column
formula <- as.formula(paste("price ~", 
          paste(c(names(train_allclasses)[-which(names(train_allclasses) %in% c("price", "class"))], 
    "I(duration^2)", "I(days_left^2)"), collapse = " + ")))
cat("Constructed formula for the model:\n")
print(formula)
```
```{r}
# Create MARS model for economy dataset with fixed parameters
mars_model_economy <- earth(formula, 
                             data = train_economy,  # Changed to use train dataset
                             degree = 4,  # Fixed degree
                            )

# Print model results
summary(mars_model_economy)
```

```{r}
par(mfrow = c(2, 2))  # Set up the plotting area
plot(mars_model_economy)
```

```{r, fig.width=10, fig.height=6}
evimp <- evimp(mars_model_economy, trim=TRUE, sqrt.=TRUE)
plot(evimp)
```

```{r}
# Predictions and performance metrics for economy model
test_economy$duration_squared <- test_economy$duration^2
test_economy$days_left_squared <- test_economy$days_left^2
predictions_economy <- predict(mars_model_economy, newdata = test_economy)

# Calculate the residuals
residuals_mars_economy <- test_economy$price - predictions_economy

# Calculate the log-likelihood (fixed array conformability issue)
log_likelihood <- -0.5 * length(residuals_mars_economy) * (log(2 * pi) + 
                  log(var(residuals_mars_economy))) - 
                  sum(residuals_mars_economy^2) / (2 * var(residuals_mars_economy))

# Calculate AIC and BIC
n <- length(residuals_mars_economy)  # Number of observations
k <- length(mars_model_economy$coefficients)  # Number of parameters

aic_value <- -2 * log_likelihood + 2 * k
bic_value <- -2 * log_likelihood + log(n) * k

# Print the results
cat("Economy Model Performance:\n")
cat("AIC:", aic_value, "\n")
cat("BIC:", bic_value, "\n")

# Performance metrics for allclasses model using test set
# R^2 for log-transformed data
r2_economy_log <- 1 - sum((residuals_mars_economy)^2) / 
  sum((test_economy$price - mean(test_economy$price))^2)

# R^2 for original scale data
r2_economy_orig <- 1 - sum((exp(test_economy$price) - exp(predictions_economy))^2) / 
                      sum((exp(test_economy$price) - mean(exp(test_economy$price)))^2)

# Calculate performance metrics
mae_economy <- mae(exp(test_economy$price), exp(predictions_economy))  
rmse_economy <- rmse(exp(test_economy$price), exp(predictions_economy))  
mape_economy <- mape(exp(test_economy$price), exp(predictions_economy))  


# Print performance results
cat("R^2 Test (log scale):", r2_economy_log, "\n")
cat("R^2 Test (original scale):", r2_economy_orig, "\n")
cat("MAE:", mae_economy, "\n")
cat("RMSE:", rmse_economy, "\n")
cat("MAPE:", mape_economy * 100, "%\n")

```
```{r}
# Create a histogram of the residuals
hist(residuals_mars_economy, breaks = 30, main = "Histogram of Residuals",
     xlab = "Residuals", ylab = "Frequency", col = "lightblue", border = "black", freq = FALSE)

# Add a density curve
lines(density(residuals_mars_economy), col = "red", lwd = 2)
```

## MARS - Business Model

```{r}
# Construct the formula for the model, excluding the 'class' column
formula <- as.formula(paste("price ~",
          paste(c(names(train_allclasses)[-which(names(train_allclasses) %in% c("price", "class"))], 
    "I(duration^2)", "I(days_left^2)"), collapse = " + ")))
cat("Constructed formula for the model:\n")
print(formula)
```

```{r}
# Create MARS model for business dataset with fixed parameters
mars_model_business <- earth(formula, 
                             data = train_business,
                             degree = 4,  # Fixed degree
                            )

# Print model results
summary(mars_model_business)
```

```{r}
par(mfrow = c(2, 2))  # Set up the plotting area
plot(mars_model_business)
```

```{r}
# Make predictions first
test_business$duration_squared <- test_business$duration^2
test_business$days_left_squared <- test_business$days_left^2
predictions_business <- predict(mars_model_business, newdata = test_business)

# Calculate the residuals
residuals_mars_business <- test_business$price - predictions_business

# Calculate the log-likelihood (fixed array conformability issue)
log_likelihood <- -0.5 * length(residuals_mars_business) * (log(2 * pi) +
                  log(var(residuals_mars_business))) - 
                  sum(residuals_mars_business^2) / (2 * var(residuals_mars_business))

# Calculate AIC and BIC
n <- length(residuals_mars_business)  # Number of observations
k <- length(mars_model_business$coefficients)  # Number of parameters

aic_value <- -2 * log_likelihood + 2 * k
bic_value <- -2 * log_likelihood + log(n) * k

# Print the results
cat("Business Model Performance:\n")
cat("AIC:", aic_value, "\n")
cat("BIC:", bic_value, "\n")

# Performance metrics for allclasses model using test set
# R^2 for log-transformed data
r2_business_log <- 1 - sum((residuals_mars_business)^2) / 
  sum((test_business$price - mean(test_business$price))^2)

# R^2 for original scale data
r2_business_orig <- 1 - sum((exp(test_business$price) - exp(predictions_business))^2) / 
                      sum((exp(test_business$price) - mean(exp(test_business$price)))^2)

mae_business <- mae(exp(test_business$price), exp(predictions_business)) 
rmse_business <- rmse(exp(test_business$price), exp(predictions_business))  
mape_business <- mape(exp(test_business$price), exp(predictions_business))  


cat("R^2 TEST (log scale):", r2_business_log, "\n")
cat("R^2 TEST (original scale):", r2_business_orig, "\n")
cat("MAE:", mae_business, "\n")
cat("RMSE:", rmse_business, "\n")
cat("MAPE:", mape_business * 100, "%\n")
```

```{r}
# Create a histogram of the residuals
hist(residuals_mars_business, breaks = 30, main = "Histogram of Residuals",
     xlab = "Residuals", ylab = "Frequency", col = "lightblue", border = "black", freq = FALSE)

# Add a density curve
lines(density(residuals_mars_business), col = "red", lwd = 2)
```

